A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period

### 2021-02-10 analysis/paper/paper.Rmd
revise Intro section to stress the method

---
title: "A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Burials provide valuable information to study social structures based on the assumption that burials and associated grave goods can represent social roles and relations in a society. To study social relationships, network analysis has been increasingly applied to archaeological data to explore network structures and patterns to infer interactions and relationships between entities. Statistical approaches to network analysis, such as exponential random graph models (ERGMs), provide a way to test hypotheses on dynamic processes of network formation. However, the computational difficulty and sensitivity to uncertainties limits the application of ERGMs. In this paper, we introduce a Bayesian framework on ERGMs that enables an efficient computational process, effective quantification of uncertainty, and robust model evaluation of network properties. We tested a hypothesis of social change relative to the arrival of Europeans by studying burial data from Kiwulan, an Iron Age site in northeastern Taiwan. The results indicate a transition from ties based on ritual objects to wealth objects, and a more centralized structure with increased social differentiation after European presence in the 17th century. Our case study demonstrates the effectiveness of Bayesian network analysis for archaeological data, and expands the use of burials in understanding the impacts of colonial presence on Indigenous groups in a pericolonial context.
keywords: |
  Bayesian modeling; network analysis; Monte-carlo simulation; Burials; European colonialism; social differentiation; Taiwan
highlights: |
  1. We present Bayesian inference on exponential random graph models (ERGMs) to model archaeological networks.
  2. We explore social processes and network structures in archaeological data from two time periods.
  3. Differences in network statistics imply change over time in social structures influenced by foreign presence. 
  4. Bayesian ERGMs are efficient and useful for evaluating and interpreting networks. 
---

@@

<!---network analysis and the basic concepts-->
Network analysis has been increasingly used by archaeologists to explore rational data to understand past relationships, interactions, or overall structures of observed phenomena [@Brandes2013; @Mills2017; @Peeples2019]. A network can be viewed as a patterned aggregation that includes individual elements, pair-wise relationships, and an overall structure showing global patterns represented in the data. Recently, network modeling provides many new tools to examine statistically the intercorrelations between individual elements in a dynamic way. Among network modeling methods, exponential random graph models (ERGMs) are one important family of statistical graphs to explore complex networks [@Harris2013; @Ghafouri2020]. However, the computational difficulties and the sensitivity to uncertainties to some extent limits its practical applications [@Caimo2014]. In this paper, we go beyond ERGMs with a novel Bayesian approach to investigate the formation of material connections using burial data from a case study in Iron Age northeastern Taiwan to understand social changes during foreign contact.

<!--study social changes in pericolonial context using burial data as a proxy-->
Social changes in Indigenous societies, when faced with colonial powers, are commonly observed in many parts of the world, especially European colonies where Indigenous economic, cultural, or socio-political aspects were substantially impacted [@Dietler2005; @Silliman2005; @Voss2005]. Recent studies demonstrate that the indirect effect of colonialism, or involvement in long-distance trade, could also have impacts on Indigenous societies, this is known as a pericolonial context [@Acabado2017; @Trabert2017; @LWandBM2020ornament]. To explore social changes, burials are important evidence to understand past societies because material culture and biological records of burial behaviors can represent the social ranking or identities of the deceased, and social relations between them [@Binford1971; @Drennan2010; @Saxe1970]. Burial treatments can further indicate social complexity or inequality where foreign goods from long-distance trade were used by Indigenous people to express status [@Carter2015; @Dolfini2019]. Usually, burial studies are based on characterizing each physical trace or burial variable, such as biological records, grave forms and goods, or ritual behaviors, and finally combining those individual observations to infer past societies [@Byrd1995; @Seikel2011]. Recent development of network modeling provides new tools to explore the structure of the material configurations and dependencies between variables to understand past relationships, interactions, and underlying mechanisms [@Brughmans2013; @Brughmans2018; @Crabtree2015; @Erik2015; @Freeman2004; @Salvini2010; @Ostborn2015].

@@

<!-- ERGMs in archaeology--> 
Exponential random graph models (ERGMs) first appeared in archaeology with @Brughmans2014, who studied Iron age settlement patterns in Southern Spain by modeling inter-settlement visibility networks and visual control at 159 sites. By fitting models with the observed network of archaeological data, they proposed that ERGMs is a promising method for exploration and hypothesis testing for social processes. Similarly, @Amati2019 modeled three networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore interaction mechanisms, including proximity, inter-cultural items, and pottery types. By comparing hypothesized networks with the observed sites, they found that the presence of hub sites can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively.

>>>>

---
title: "A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Burials provide valuable information to study social structures based on the assumption that burials and associated grave goods can represent social roles and relations in a society. To study social relationships, network analysis has been increasingly applied to archaeological data to explore network structures and patterns to infer interactions and relationships between entities. Statistical approaches to network analysis, such as exponential random graph models (ERGMs), provide a way to test hypotheses on dynamic processes of network formation. However, the computational difficulty and sensitivity to uncertainties limits the application of ERGMs. In this paper, we introduce a Bayesian framework on ERGMs that enables an efficient computational process, effective quantification of uncertainty, and robust model evaluation of network properties. We tested a hypothesis of social change relative to the arrival of Europeans by studying burial data from Kiwulan, an Iron Age site in northeastern Taiwan. The results indicate a transition in the process that led to the formation of the burial network, from ties based on ritual objects to wealth objects, and a more centralized structure with increased social differentiation after European presence in the 17th century. Our case study demonstrates the effectiveness of Bayesian network analysis for archaeological data, and expands the use of burials in understanding the impacts of colonial presence on Indigenous groups in a pericolonial context.
keywords: |
  Bayesian modeling; network analysis; Monte-carlo simulation; Burials; European colonialism; social differentiation; Taiwan
highlights: |
  1. We present Bayesian inference on exponential random graph models (ERGMs) to model archaeological networks.
  2. We explore social processes and network structures in archaeological data from two time periods.
  3. Differences in network statistics imply change over time in social structures influenced by foreign presence. 
  4. Bayesian ERGMs are efficient and useful for evaluating and interpreting networks. 
---

@@

<!---network analysis and the basic concepts-->
Network analysis has been increasingly used by archaeologists to understand past relationships, interactions, and structures of observed phenomena by visualizing and analyzing rational data [@Brandes2013; @Mills2017; @Peeples2019]. The recent development of modeling approaches enable hypothesis testing for the intercorrelations between individual elements and overall structures inherit in networks [@Brughmans2013; @Brughmans2018; @Freeman2004; @Salvini2010]. By comparing hypothesized networks with the observed network, network modeling can answer a wide range of anthropological questions related to exchange [@Gjesfjeld2013; @Crabtree2015], diffusion [@Ostborn2015], or social transformation [@Mills2013] in a statistical way. Among network modeling methods, exponential random graph models (ERGMs) are stochastic models for investigating the process of network formation through dependence assumptions for relationships and simulating network patterns [@Harris2013; @Ghafouri2020]. Since relationships emerging relative to their position in a network can be hypothesized, ERGMs allows evaluation of dynamic social processes behind observed archaeological networks [@Brughmans2014]. Although ERGMs has potential to study social processes, computational difficulties and the sensitivity to uncertainties to some extent limits the practical applications of ERGMs, where results could be challenging for interpretation [@Caimo2014].

<!--study social changes in pericolonial context using burial data as a proxy-->
In this paper, we go beyond ERGMs with a novel Bayesian approach that can alleviate the computational issues in ERGMs to better interpret the process of network formation for past phenomena. We use Baysien ERGMs to explore social changes by investigating the formation of material connections between burials from a case study in Iron Age northeastern Taiwan. Social changes in Indigenous societies, when faced with colonial powers, are commonly observed in many parts of the world, especially European colonies where Indigenous economic, cultural, or socio-political aspects were substantially impacted [@Dietler2005; @Silliman2005; @Voss2005]. Recent studies demonstrate that the indirect effect of colonialism, or involvement in long-distance trade, could also have impacts on Indigenous societies, this is known as a pericolonial context [@Acabado2017; @Trabert2017; @LWandBM2020ornament]. To explore social changes, burials are important evidence to understand past societies because material culture and biological records of burial behaviors can represent the social ranking or identities of the deceased, and social relations between them [@Binford1971; @Drennan2010; @Saxe1970]. Burial treatments can further indicate social complexity or inequality where foreign goods from long-distance trade were used by Indigenous people to express status [@Carter2015; @Dolfini2019].

Usually, burial studies are based on characterizing each physical trace or burial variable, such as biological records, grave forms and goods, or ritual behaviors, and finally combining those individual observations to infer past societies [@Byrd1995; @Seikel2011]. We treat burials as a complex network based on the assumption that a network is a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data [@Brandes2013].

@@

<!-- ERGMs in archaeology-->

Exponential random graph models (ERGMs) first appeared in archaeology with @Brughmans2014, who studied Iron age settlement patterns in Southern Spain by modeling inter-settlement visibility networks and visual control at 159 sites. By fitting models with the observed network of archaeological data, they proposed that ERGMs is a promising method for exploration and hypothesis testing for social processes. Similarly, @Amati2019 modeled three networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore interaction mechanisms, including proximity, inter-cultural items, and pottery types. By comparing hypothesized networks with the observed sites, they found that the presence of hub sites can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively.
